Effects of COVID-19 related social media use on well-being

### 2023-08-24 manuscript.Rmd
- shortened manuscript for submission

floatsintext      : yes
endnotes          : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021].
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022].
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children.
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question
[e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021],
<!-- [@bendauAssociationsCOVID19Related2021; @bradleyStressMoodSmartphone2021; @choiMediatedCommunicationMatters2021; @dornemannHowGoodBad2021; @edenMediaCopingCOVID192020; @guazziniSecondWaveAnalysis2022; @latikkaLonelinessPsychologicalDistress2022; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020; @sewallObjectivelyMeasuredDigital2021; @stainbackCOVID1924News2020; @yuePassiveSocialMedia2022] -->
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

<!-- Two underlying theories guided the selection of variables for this study, namely the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being. -->
<!-- <!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
<!-- Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018]. -->
<!-- Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning. -->
<!-- Subjective well-being is primarily about achieving positive and avoiding negative affect. -->
<!-- One of the most prominent indicators of well-being is life satisfaction. -->
<!-- In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept combining psychological and subjective well-being. -->
<!-- Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020]. -->
<!-- <!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
<!-- To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect. -->
<!-- Together, this should provide an encompassing perspective on potential media effects. -->

<!-- The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology. -->
<!-- First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_. -->
<!-- Distinguishing different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types. -->
<!-- Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is often considered more negative [@dienlinImpactDigitalTechnology2020]. -->
<!-- Similarly, branded apps are separate communication entities with potentially divergent effects and affordances. -->
<!-- For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. -->
<!-- Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical. -->
<!-- First studies suggest that during the pandemic Instagram use was indeed more beneficial for well-being than Facebook use [@masciantonioDonPutAll2021]. -->
<!-- In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective. -->
<!-- Together, this should offer a nuanced and comprehensive understanding of communication. -->

This study investigates how different _facets_ of well-being are affected by different _types_ and different _channels_ of communication [@meierComputermediatedCommunicationSocial2020a].
<!-- , namely active and passive use [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a]. -->
<!-- Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a]. -->
<!-- Reading is generally considered as passive and writing as active use, while there are also specific behaviors falling somewhere in-between such as liking or sharing content [@meierDoesPassiveSocial2022]. -->
<!-- In this study,  -->
Building on the typology of subjective well-being [@dienerAdvancesOpenQuestions2018], three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used [@verduynPassiveFacebookUsage2015], I further distinguish three types of use and five popular channels. 
The types of use include reading, liking and sharing, and posting COVID-19 related content.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting and liking COVID-19 related news.
Liking and sharing are combined as they both represent low-threshold, platform-ingrained, easily quantifiable interactions.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

<!-- <!-- In conclusion, most effects are likely somewhere between trivial and small. -->
<!-- <!-- I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small. -->

<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are short-term, with well-being after some time returning to prior levels [@dienerAdvancesOpenQuestions2018].
Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
So although well-being can be affected by external events and factors, this does not happen easily.
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.
<!-- For example, using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022]. -->
<!-- INCLUDE IF SPACE -->
<!-- On the other hand, although mood management theory suggests that effects should not be overly negative, it could be that although short-term effects are positive, long-term effects are negative.  -->
<!-- Precisely because social media have so many positive consequences in the short run, this might cause problem in the long run---for example, because social media use displaces other potentially more meaningful and/or relaxing activities [@hallSocialMediaUse2022]. -->
<!-- As with many other things, there can be too much of a good thing. -->
<!-- It is therefore often asked whether social media can become addictive, with users sometimes expressing this fear themselves. -->
<!-- END -->
<!-- it is still possible that people misjudge media effects. -->
<!-- Users are often overly optimistic [@metzgerComparativeOptimismPrivacy2017], focusing on short term benefits while ignoring long-term costs.  -->
<!-- In other words, social media might not problematic because they are inherently bad, but rather because they are too good. -->
<!-- However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020]. -->
<!-- In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media use represent a new regular behavior that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018]. -->
<!-- In conclusion, there are several plausible theoretical mechanisms as to why effects might be negative or positive, explaining why on average effects might not be particularly negative or positive. -->

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
<!-- , outlining mechanisms of potential positive effects.  -->
<!-- At the same time, there are also several plausible reasons as to why COVID-19 related social media use is beneficial for well-being. -->
<!-- Access to information and resources:  -->
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
<!-- It allows individuals to stay informed about the latest updates, guidelines, and recommendations from reputable health organizations and experts.  -->
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
<!-- Educational opportunities:  -->
<!-- Social media can be a valuable _educational tool_, with various organizations and experts sharing informative content about COVID-19 [@worldhealthorganizationCoronavirusDiseaseCOVID192023].  -->
<!-- <!-- This can include explanatory videos, infographics, and articles that help individuals understand the virus, its transmission, prevention strategies, and vaccination information.  -->
<!-- Access to educational resources on social media can empower individuals to make informed decisions regarding their health. -->
<!-- Community support and solidarity:  -->
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
<!-- Sharing experiences, offering support, and expressing solidarity can foster a sense of community and reduce feelings of isolation.  -->
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
<!-- Promoting mental health and well-being:  -->
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic [@twitterTalkingMentalHealthAwarenessTwitter2020]. 
<!-- They provide guidance on coping mechanisms, stress reduction techniques, and self-care practices.  -->
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
<!-- Encouraging positive behaviors:  -->
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@atheyDigitalPublicHealth2023; @huntSocialMediaBased2022]. 
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.
<!-- Amplifying diverse voices and perspectives:  -->
<!-- Social media platforms provide a space for diverse voices and perspectives to be heard [@brunsAreFilterBubbles2019].  -->
<!-- This allows individuals to access a range of viewpoints, experiences, and cultural practices related to COVID-19.  -->
<!-- Exposure to diverse perspectives might promote empathy, understanding, and a more comprehensive understanding of the pandemic's impact on different communities. -->

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
<!-- Misinformation and rumors:  -->
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
<!-- Overwhelm and anxiety:  -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022].
<!-- Confirmation bias and echo chambers:  -->
<!-- Social media algorithms are designed to show users content that aligns with their interests and beliefs, potentially contributing to _echo chambers_ [@cinelliEchoChamberEffect2021].  -->
<!-- <!-- This might lead to confirmation bias, where individuals are more likely to be exposed to information that confirms their preexisting beliefs.  --> -->
<!-- In the context of COVID-19, this might reinforce false or misleading information and prevent individuals from considering alternative perspectives. -->
<!-- Cyberbullying and harassment:  -->
Social media platforms are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts. 
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
<!-- Social comparison and fear of missing out (FOMO):  -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].
<!-- Reduced face-to-face interaction:  -->
<!-- Spending excessive time on social media to learn about COVID-19 can lead to a decrease in face-to-face interactions and real-life connections [but see @hallSocialMediaUse2022].  -->
<!-- This might impact mental health, as most humans have an innate need for social support and physical presence, which online interactions might not fully replicate. -->

<!-- Using Social media is ambivalent. -->
<!-- It can impair well-being when causing embarrassment, stress, or disinformation, and it can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021]. -->
<!-- Although social media are often associated with negative outcomes, there are a large number of positive effects, which either offset or at least significantly reduce negative outcomes. -->
<!-- Positive outcomes encompass finding relevant information; maintaining and fostering relationships; expressing one's personality; or entertaining oneself [e.g., @pelletierOneSizeDoesn2020]. -->
<!-- Negative effects include distraction; displacement of other meaningful activities; consumption of shallow, misleading or challenging content; or negative social comparisons [e.g., @meierDoesPassiveSocial2022]. -->
<!-- People are much more likely to use media that offer plentiful benefits [@katzUsesGratificationsResearch1973]. -->
<!-- Users implicitly learn what media help them regulate their mood and thereby well-being [@zillmannMoodManagementCommunication1988]. -->
<!-- Given the ubiquitous use of social media, the many benefits they offer, plus the general and implicit competence to use media that foster mood management, it is hence unlikely that the effects of social media use on well-being are profusely negative. -->

<!-- As outline above,   -->
There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@stainbackCOVID1924News2020; @yuePassiveSocialMedia2022]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, engaging more on social media was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
Several studies reported mostly neutral effects of social media use on well-being indicators [@edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

<!-- ## Empirical Studies on Social Media Effects -->

<!-- There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically. -->
<!-- Results are mixed. -->
<!-- When browsing social media for COVID-19 related news, many users reported being captivated to such an extent they could not stop using social media [@kleinDarklySoothingCompulsion2021]. -->
<!-- During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021], for example to cope with the required physical distancing measures. -->
<!-- In a study with 1,131 residents from Wuhan in China, people who spent more time in quarantine also spent more time on social media [@yuePassiveSocialMedia2022].  -->
<!-- Those who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress. -->
<!-- People who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [...] and significantly more specific COVID-19 related anxiety symptoms" [@bendauAssociationsCOVID19Related2021, p. 288]. -->
<!-- @edenMediaCopingCOVID192020 analyzed how 425 US college students used media during the first wave of the pandemic, finding both positive and negative relations with well-being. -->
<!-- In a sample of 312 respondents collected via Amazon Mechanical Turk, people who used media to attain information were more lonely and less satisfied with their lives [@choiMediatedCommunicationMatters2021]. -->
<!-- In a large-scale study with 11,537 respondents from the US, increased COVID-19-related media consumption was related to more psychological distress [@stainbackCOVID1924News2020]. -->
<!-- A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use during the pandemic on mental health, finding that digital technology did not have significant effects on mental health [@sewallObjectivelyMeasuredDigital2021; for a similar study with comparable results, see @bradleyStressMoodSmartphone2021]. -->
<!-- A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness increased [@guazziniSecondWaveAnalysis2022]. -->
<!-- In a study with 735 participants from Finland, levels of loneliness were stable during the pandemic, and people who engaged more on social media experienced less loneliness [@latikkaLonelinessPsychologicalDistress2022]. -->
<!-- Together, the literature is mixed, showing both positive and negative affects of social media use focused on or during COVID-19 on well-being [see also @dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020]. -->

@@

<!-- Although the average effect of use on well-being is likely very small, it might be that specific subgroups exhibit larger effects. -->

<!-- > Research question: How do the effects of COVID-19 related social media use on well-being depend on gender, age, immigration background, education, and urban living context? -->

<!-- Because media effects can differ across users, situations, and content [@livingstoneEuropeanResearchChildren2018; @valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically. -->
<!-- First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating. -->
<!-- Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide. -->
<!-- If one learns about such events, the initial reaction might be shock, fear, or dismay. -->
<!-- Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life. -->
<!-- <!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
<!-- That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021]. -->
<!-- A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased [@guazziniSecondWaveAnalysis2022]. -->

<!-- In a study with 735 participants from Finland, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media experienced less loneliness [@latikkaLonelinessPsychologicalDistress2022]. -->
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

<!-- Second, there could also be _indirect_ effects. -->
<!-- Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017]. -->
<!-- To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising. -->
<!-- Another potentially negative mechanism at play are problematic social comparison processes.  -->
<!-- Therefore, if a society collectively overuses social media during a pandemic, there is potential for negative effects. -->
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

<!-- On the other hand, one can make the case that using social media for COVID-19 related reasons might even be beneficial, especially in times of a pandemic. -->
<!-- Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time. -->
<!-- In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea. -->
<!-- Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments. -->
<!-- <!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

<!-- To summarize, it seems that from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible.  -->
<!-- Building on established theories from Communication and current empirical findings, we would not assume that effects are either profoundly negative or strongly positive. -->

<!-- However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below). -->
<!-- We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education. -->

<!-- The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched. -->
<!-- This also holds true for the different types of communication such as active or passive use. -->
<!-- What determines whether or not an effect should be considered small or trivial? -->
<!-- As a starting point, we could refer to standardized effect sizes. -->
<!-- According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10. -->
<!-- Finally, several individual studies employing advanced methods found smalls relations between social media use and well-being [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020]. -->
<!-- Although for roughly one quarter of all users the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral [@beyensSocialMediaUse2021]. -->
<!-- This finding is aligned with the Differential Susceptibility to Media Effects Model:  -->
<!-- Although there is substantial _variation_ of media effects for individual users, the _average_ effects reported in the literature are often small [@valkenburgDifferentialSusceptibilityMedia2013].  -->

<!-- Together, the strongest argument to me is that _in general_ the effects of social media on well-being are, on average, small at best.  -->
<!-- Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further. -->
<!-- Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear. -->
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->

<!-- A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small, potentially too small to matter [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- And indeed, several if not most of the current meta-analyses find small to trivial effects [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021]. -->

<!-- In this study, this general hypothesis will be analyzed specifically for the three communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content. -->
<!-- In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.  -->

@@

<!-- # Current Study -->
## Smallest Effect Size of Interest
<!-- ## Inference Criteria -->

Testing the hypothesis necessitates defining what is considered a "trivial effect size". 
<!-- Effectively testing this hypothesis  -->
<!-- Above I already referred to standardized effect sizes. -->
<!-- However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009]. -->
<!-- Standardized effect sizes are determined by a sample's variance,[^cohensd] which is problematic:  -->
<!-- The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected. -->
<!-- Instead, it should depend on absolute criteria. -->
To this end, we need to define a so-called smallest effect size of interest (SESOI) [@lakensEquivalenceTestingPsychological2018].
A trivial effect would then need to be smaller than the SESOI (see below).
What could be a minimally interesting, nontrivial effect?
Being a normative question, finding a clear, single, or unanimous answer is impossible.
<!-- In the end, it is a personal question. -->
However, it is still necessary and helpful to work toward a plausible benchmark. 
I suggest the following SESOI for this research question:

>>>>

floatsintext      : yes
endnotes          : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : yes
draft             : no

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021].
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022].
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children.
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question
[e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021],
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

This study investigates how different _facets_ of well-being are affected by different _types_ and different _channels_ of communication [@meierComputermediatedCommunicationSocial2020a].
Building on the typology of subjective well-being [@dienerAdvancesOpenQuestions2018], three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used [@verduynPassiveFacebookUsage2015], I further distinguish three types of use and five popular channels. 
The types of use include reading, liking and sharing, and posting COVID-19 related content.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting and liking COVID-19 related news.
Liking and sharing are combined as they both represent low-threshold, platform-ingrained, easily quantifiable interactions.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are short-term, with well-being after some time returning to prior levels [@dienerAdvancesOpenQuestions2018].
Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
So although well-being can be affected by external events and factors, this does not happen easily.

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic [@twitterTalkingMentalHealthAwarenessTwitter2020]. 
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@atheyDigitalPublicHealth2023; @huntSocialMediaBased2022]. 
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022].
Social media platforms are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts. 
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@stainbackCOVID1924News2020; @yuePassiveSocialMedia2022]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, engaging more on social media was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
Several studies reported mostly neutral effects of social media use on well-being indicators [@edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

@@

## Smallest Effect Size of Interest

Exploring the hypothesis entails establishing the criteria for discerning a 'trivial effect size.' 
To achieve this, it is necessary to define the smallest effect size of interest (SESOI) [@anvariUsingAnchorbasedMethods2021a]. 
In this context, a trivial effect should fall below the threshold set by the SESOI (detailed below).
Determining what constitutes a minimally intriguing, nontrivial effect is a matter with normative implications, making it challenging to arrive at a definitive, singular consensus. 
Nonetheless, it remains both essential and beneficial to strive for a reasonable benchmark. 
With that in mind, I propose the following SESOI as a suitable reference point for this study:

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
According to @normanInterpretationChangesHealthrelated2003, people can reliably notice seven levels of change in satisfaction with health.
So if satisfaction is measured on a 7-point scale, a four unit change in social media use should result in a one unit change in life satisfaction. 
In this study, life satisfaction was measured on an 11-point scale, and affect on a 5-point scale. 
Transposed to this scaling, the SESOI for life satisfaction is _b_ =  ±.30, and for positive and negative affect _b_ = ±.15 (for more information, see online supplementary material).

@@

## Data Analysis

The hypothesis is analyzed using the interval testing approach as proposed by @dienesUsingBayesGet2014.
To illustrate, let us consider the case of life satisfaction [SESOI: ±.30]. 
Here, the null-region lies between -.30 and +.30.
If the 95% confidence interval falls completely within the null-region (e.g., _b_ = -.05, [95% CI: -.15, .05]), the hypothesis that the effect is trivial is supported.
If the confidence interval falls completely outside of the null-region (e.g., _b_ = -.40, [95% CI: -.45, -.35]), the hypothesis is rejected and the existence of a meaningful negative effect is supported.
If the confidence interval and the null region overlap (e.g., _b_ = -.30, [95% CI: -.35, -.25]), the hypothesis is not supported and the results are considered inconclusive, while a meaningful positive effect is rejected.
If the confidence interval exceeds both sides of the null region (e.g., _b_ = -.025, [95% CI: -.40, .35]), the hypothesis is not supported and judgement is suspended.
For an illustration, see Figure \@ref(fig:sesoi).
